Covariate analysis of UFL-Mayo-ISB RNASeq from cerebellum

### 2016-01-25 code/Rmd/Mayo_Cerebellum_RNASeq.Rmd
Minor changes

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA <- METADATA %>%
  dplyr::filter(!(SampleID %in% c("742_CER", "11386_CER","11311_CER","1923_CER","1147_CER")))
```
Following five samples were considered outliers and removed 742_CER, 11386_CER, 11311_CER, 1923_CER, 1147_CER
```{r preprocess.data1,cache=TRUE}
indToRetain = intersect(METADATA$SampleID, colnames(COUNT))

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(Diagnosis), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") 
p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

@@

WGCNA::plotDendroAndColors(sampleTrees, 
                    labels2colors(cbind(tree,COVARIATES)),
                    groupLabels = c('Cluster',colnames(COVARIATES)))
collectGarbage()
```

@@

# Linear model with new weights and design
VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION$E, 
                          design=DM1$design, 
                          weights = VOOM.ADJUSTED.GENE_EXPRESSION$weights)  
```
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r calculate.residuals, include=FALSE}
# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                              VOOM.ADJUSTED.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              DESIGN,
                                              'residual matrix of all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
```{r addVarsBack}
varsToAddBackIn <- c("DiagnosisAD", "DiagnosisControl", "DiagnosisPSP", "DiagnosisPathologic Aging")
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(VOOM.ADJUSTED.FIT$coefficients[,varsToAddBackIn]))
RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'ensembl_gene_id') 
```

```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = as.data.frame(summary(COVARIATES$Diagnosis))
colnames(tmp) = 'No. of samples'
kable(tmp)
```

>>>>

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA <- METADATA %>%
  dplyr::filter(!(SampleID %in% c("742_CER", "11386_CER","11311_CER","1923_CER","11477_CER")))
```
Following five samples were considered outliers and removed 742\_CER, 11386\_CER, 11311\_CER, 1923\_CER, 11477\_CER
```{r preprocess.data1}
indToRetain = intersect(METADATA$SampleID, colnames(COUNT))

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(Diagnosis), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

@@

WGCNA::plotDendroAndColors(sampleTrees, 
                    labels2colors(cbind(tree,COVARIATES)),
                    groupLabels = c('Cluster',colnames(COVARIATES)))
collectGarbage()
dev.off()
```

@@

# Linear model with new weights and design
VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION$E, 
                          design=DM1$design, 
                          weights = VOOM.ADJUSTED.GENE_EXPRESSION$weights)  
```
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r calculate.residuals, include=FALSE, results = 'asis'}
# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                              VOOM.ADJUSTED.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              DESIGN,
                                              'residual matrix of all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Add "Diagnosis"" back to the residuals
```{r addVarsBack}
varsToAddBackIn <- c("DiagnosisAD", "DiagnosisControl", "DiagnosisPSP", "DiagnosisPathologic Aging")
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(VOOM.ADJUSTED.FIT$coefficients[,varsToAddBackIn]))
RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'ensembl_gene_id') 
```

### Clustering residual gene expression
```{r decompse.normalise.data1, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1)], scale.=T)

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(Diagnosis), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

# Tree based clustering
sampleTrees = flashClust::hclust(dist(t(RESIDUAL.GENE_EXPRESSION[,-(1)])), method = "average")
tree = WGCNA::cutreeStatic(sampleTrees, cutHeight = 200, minSize=3)

WGCNA::plotDendroAndColors(sampleTrees, 
                           labels2colors(cbind(tree,COVARIATES)),
                           groupLabels = c('Cluster',colnames(COVARIATES)))
collectGarbage()
dev.off()
```

```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = data.frame(summary(COVARIATES$Diagnosis))
colnames(tmp) = "No. of samples"
kable(tmp)
```
